Estimating Gaze Duration Error from Eye Tracking Data

### 2022-10-19 paper/paper.tex
paper updates

Gaze fixation models will typically have an error profile that is dependent on both 
the nature of the media being presented, the variety of faces and lighting conditions 
in the training data. In addition, it has been observed that the error profile varies 
depending on the location of the true fixation point within the subject's field of view.
The errors in fixation point measurement for a given eye tracking solution are routinely
monitored and adjusted through a process called calibration, in which users are presented
a series or \emph{required fixation locations} (RFL) which are compared with measured 
points of fixation\cite{Hornof2002}. 
Improvements in the callibration process
are an ongoing focus for the development of algorithms\cite{Zhang2014,Hassoumi2019} 
and software tools\cite{ETCAL2018}. In some scenarios there are additional considerations
such as whether to use fixation data from individual eyes, or a composite 
signal\cite{Hooge2019}.

Typically the nature of the error detected in the calibration process can be broken
into both accuracy and precision. Wherein accuracy is the tendency for the measure
fixation points to be closer to true fixation, and precision is the tendency for the
measured locations to vary for a given true fixation\cite{Holmqvist2022}. This distinction
is critical because it forms the basis of methods that can correct for the kind of 
systematic bias observed with low precision eye tracking. It also been observed that
metrics of the precision in these models vary in the way they rank technologies\cite{Niehorster2020}. We can conclude from this that the patterns of systematic error in eye tracking
data are more complicated that can be quanitifed by a single metric.

@@

The errors of eye tracking data technology can be decomposed into a range of
independent sources that researchers need to consider \cite{Holmqvist2012}.
Unless they are properly addressed these sources of error can manifest themselves
as systematic biases across undesirable dimensions such as the age\cite{Dalrymple2018}
or ethnicity of subjects\cite{Blignaut2013}.

@@

Formally stated this is the probability distribution over measured durations $\hat{d}$ given a
known duration $d$, the session length $s$ and the gaze target location $l$. 
We express this distribution as $P(\hat{d}|d,l,s)$. The
procedure for making this estimation involves iteratively generating random samples
of gaze fixation paths $\mathcal{F}$ consistent with the parameters $d$, $s$ and $l$. 
For each point $f$ in $\mathcal{F}$ we determine a measurement point $\hat{f}$ 
by drawing samples from the eye tracking calibration set. 
The samples from the calibration data are used to determine the $\delta_x$ and $\delta_y$ 
error in the measurement of gaze fixation point $f$. 
We draw these delta values from the calibration file such the
the probability of the point being chosen is propotional to its distance from the point $f$
in the generated path. The complete algorithm
for estimating $P(\hat{d}|d,l,s)$ is shown in Algorithm \ref{alg:p_of_dhat}

@@

The bulk of the work is done by two functions inside the simulation loop. The first function
$generatePath(d,l,s)$ takes a true gaze duration (for that simulation iteration) the location
of the gaze target and the viewing session length. It then
generates a random path of fixation points that is compatible with those parameters.
The second function 
$generateMeasurement(p,l,C)$, takes the path, target location and the eye tracking calibration 
data and produces a sample of measured attention time. It does this by using the calibration data to sample
fixation errors and apply them to the path data. The noisy path data is then used to determine
the estimated gaze duration by looking at the number of fixation points that intersect with the target
location.
Repeated application of this process for a given value of $d$ produces multiple samples of $\hat{d}$
for that true gaze duration. Repeating the process over different 
values of $d$ allows us to populate a two dimensional array where the first dimension is the
true duration and the second the estimated duration. Note that our probability distribution is discrete
because we restrict ourselves to sampling over a fixed set of increments 
between zero and the session length.

@@

In real world applications we will have the true session length but rely on the model for the 
estimate of gaze duration. Meaning that the probability distribution we want is the distribution 
over true gaze durations for a given measurement, rather than the inverse which is what we 
have produced in the previous section. We produce an estimate of $P(d|\hat{d},l,s)$
through an application of Bayes' rule, as shown in Equation \ref{eq:p_of_d}.

@@

We use a uniform prior for $P(d)$, meaning in the absence of additional information all
gaze durations less than the session length are equally likely. As our distributions are
discrete estimations of an underlying continuous distribution the value of $P(d)$ is
equal to $1/(1+floor(10*s))$
 We can estimate the value
of the denominator $P(\hat{d},l,s)$ by iterating over all values of d and summing the product
of $ P(\hat{d},l,s|d) \dot P(d)$. This fully explicated form is shown in 
Equation \ref{eq:p_of_d_full}.

@@

For the sake of our simulations we use a fixed sized simulated device with viewport pixel 
dimensions of 350 by 627. Our area of interest (AOI) is a fixed position medium rectangle ad unit (MREC)
that is located toward the top of the screen. We iterate over a range of potential mean error
values (determined to be the mean Euclidean distance between the measured fixation and actual fixation).
For each of these mean error values we generate a synthetic eye tracking calibration file. We produce
two versions of this calibration data, one in which the error is distributed around the true target,
and a second biased version in which the error tends toward the upper left of the true position. We 
include the biased calibration data as it is consistent with our observations of real eye tracking data
and we explore the extent to which this bias affects gaze duration measurements.

@@

The left hand plot shows the complete error surface in which you will observe large spikes in
error in the corner correspinding to high MAE for the fixation models and high measurements of
gaze duration. This effect is caused by a complete absence of probability density in these
regions, meaning that the error is maximal because the expected value is zero. This effect 
was consistenlt observed for 
eye tracking models with high fixation error and when the measured duration was high. Meaning
that our model estimates that when fixation error is this high, measurements of these values
become extremely unlikley to be observed.

We added the right hand plot in Figure \ref{fig:error_surface} to illustrate the error surface
in the absence of these deceptive outliers. We see that in general the expected error in a gaze
duration is maximal around the mid point of the measurement range. The effect is lessened by 
lower MAE in the fixation models, but remains consistent for all levels of fixation error.

@@

We have demonstrated that the statistial information contained in eye tracking validation
data can be utilised for estimating errors in gaze duration estimation. The process involves
simulating the distribution of measured duration for each possible true duration, 
and then inverting it with Bayes rule into a distribution over true duration for a given measurement.

@@

Finally, our simualtions revealed that for large error in fixation points certain measurements
become so unlikely that our monte carlo simualtions placed none of the probability density
in those regions. This strongly suggests that an additional downside to large fixation point error
is a reduced range of measurement for area of interest studies.

>>>>

The source of errors in eye tracking technology can be decomposed into a range of
independent sources that researchers need to consider\cite{Holmqvist2012}.
Unless they are properly addressed these sources of error can manifest themselves
as systematic biases across undesirable dimensions such as the age\cite{Dalrymple2018}
or ethnicity of subjects\cite{Blignaut2013}. Machine learning based eye tracking solutions
can be impacted by the variety of faces and lighting conditions in the training data. 
The errors in fixation point measurement for a given eye tracking solution are routinely
monitored and adjusted through a process called calibration, in which users are presented
a series or \emph{required fixation locations} (RFL) which are compared with measured 
points of fixation\cite{Hornof2002}. 
Improvements in the callibration process
are an ongoing focus for the development of algorithms\cite{Zhang2014,Hassoumi2019} 
and software tools\cite{ETCAL2018}. In some scenarios there are additional considerations
such as whether to use fixation data from individual eyes, or a composite 
signal\cite{Hooge2019}.

In eye tracking literature the error detected in the calibration process is classified
using two independent metrics; accuracy and precision. Accuracy is a measure for how
close, on average, the fixation points are to the true point of fixation.
Precision is the tendency for the measured locations to vary between subsequent measurements
for a given point of fixation\cite{Holmqvist2022}. This distinction
is critical because it forms the basis of methods that can correct for the kind of 
systematic bias observed with low precision eye tracking. It also been observed that
metrics of the precision in these models vary in the way they rank 
technologies\cite{Niehorster2020}. We can conclude from this that the patterns of 
systematic error in eye tracking
data are more complicated that can be quantifed by a single metric.

@@

@@

Formally stated this is the probability distribution over measured durations $\hat{d}$ given a
known duration $d$, the session length $s$ and the gaze target location $l$. 
We express this distribution as $P(\hat{d}|d,l,s)$. The
procedure for making this estimation involves iteratively generating random samples
of gaze fixation paths $\mathcal{F}$ consistent with the parameters $d$, $s$ and $l$. 
For each point $f$ in $\mathcal{F}$ we determine a measurement point $\hat{f}$ 
by drawing samples from the eye tracking calibration set.

The samples from the calibration data are drawn as a delta in the two dimensional Cartesian
space of the field of view. Each delta consists of $\delta_x$ and $\delta_y$ which correspond
to the error in the $x$ and $y$ dimensions. We apply these delta values to the true gaze 
fixation point $f$ to create the simulated measurement $\hat{f}$.
We draw these delta values from the calibration file such the
the probability of choosing a delta is propotional to the distance between the point $f$
in the generated path, and the true fixation point in the callibration set.

The complete algorithm for estimating $P(\hat{d}|d,l,s)$ is shown in Algorithm \ref{alg:p_of_dhat}

@@

The bulk of the work is done by two functions inside the simulation loop. The first function
$generatePath(d,l,s)$ takes a true gaze duration (for that simulation iteration) the location
of the gaze target and the viewing session length. It then
generates a random path of fixation points that is compatible with those parameters.
The second function 
$generateMeasurement(p,l,C)$, takes the path, target location and the eye tracking calibration 
data and produces a sample of measured attention time. It does this by using the calibration data to sample
fixation errors and apply them to the path data. The noisy path data is then used to determine
the estimated gaze duration by looking at the number of fixation points that intersect with the target
location.
Repeated application of this process for a given value of $d$ produces multiple samples of $\hat{d}$
for that true gaze duration. Repeating the process over different 
values of $d$ allows us to populate a two dimensional array where the first dimension is the
true duration and the second the estimated duration. Note that our probability distribution is discrete
because we restrict ourselves to sampling over a set of fixed time period increments ($i$) 
between zero and the session length.

@@

In real world applications we will have the true session length but rely on the model for the 
estimate of gaze duration. Meaning that the probability distribution we want is the distribution 
over true gaze durations for a given measurement. In the previous section we estimated the inverse of
this: the distribution of measurements of a given true duration.

In order to estimate the desired distribution $P(d|\hat{d},l,s)$ we need to use 
Bayes' rule, as shown in Equation \ref{eq:p_of_d}.

@@

We use a uniform prior for $P(d)$, meaning in the absence of additional information all
gaze durations less than the session length are equally likely. As our distributions are
discrete estimations of an underlying continuous distribution the value of $P(d)$ is
equal to $1/(1+floor(i*s))$. We can estimate the value
of the denominator $P(\hat{d},l,s)$ by iterating over all values of $d$ and summing the product
of $ P(\hat{d},l,s|d) \dot P(d)$. This fully explicated form is shown in 
Equation \ref{eq:p_of_d_full}.

@@

For the sake of our simulations we use a simulated device with viewport pixel 
dimensions of 350 by 627. Our area of interest (AOI) is a fixed position medium rectangle ad unit (MREC)
that is located toward the top of the screen. We iterate over a range of potential mean error
values (equivalent to varying the accuracy), determined to by the mean Euclidean distance 
between the measured fixation point and actual fixation points.
For each of these mean error values we generate a synthetic eye tracking calibration file. We produce
two versions of this calibration data, one in which the error is centered around the true target,
and a second biased version in which the error tends toward the upper left of the true position. We 
include the biased calibration data as it is consistent with our observations of real eye tracking data
and we explore the extent to which this bias affects gaze duration measurements.

@@

The left hand plot shows the complete error surface in which you will observe large spikes in
error in the corner corresponding to both high mean error for the fixation models and 
high measurements of gaze duration. 
This effect is caused by a complete absence of probability density in these
regions, meaning that the error is maximal because the expected value is zero. This effect 
was consistently observed for 
eye tracking models with high mean fixation error and when the measured duration was high. 
The consequence of this is that our model estimates that when fixation error is this high, 
measurements of these values
become extremely unlikley to be observed.

We added the right hand plot in Figure \ref{fig:error_surface} to illustrate the error surface
in the absence of these deceptive outliers. We see that, in general, the expected error in a gaze
duration measurement is maximal around the mid point of the measurement range. The effect is lessened by 
lower mean error in the fixation models, but remains consistent for all levels of fixation error.

@@

We have demonstrated that the statistial information contained in eye tracking calibration
data can be utilised for estimating errors in gaze duration estimation. The process involves
simulating the distribution of measured duration for each possible true duration, 
and then inverting it with Bayes rule into a distribution over true duration for a given measurement.

@@

Finally, our simulations revealed that for large error in fixation points, certain measurements
become so unlikely that our Monte Carlo simualtions placed none of the probability density
in those regions. This strongly suggests that an additional downside to large fixation point error
is a reduced range of measurement for area of interest studies.
